State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-09-17 analysis/paper/paper.Rmd
re-knit all 3 versions

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
- AckermanLibrary.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

Ryan and Kevin decided to not go down this path. One quick comparison is total escapement vs. window counts; which shows up in one of our tables, and we discuss it briefly in the results. -->

@@

```{r stadem-examp-fig, fig.width=3.5, fig.height=7, fig.cap = "Schematic of how the STADEM model works. (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the re-ascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportions by origin (D), to estimate unique fish by origin."}

@@

# choose appropriate color scale for color vs. black & white
p1 = ggplot(data.frame(x = c(-5, 8)),
            aes(x)) +
  stat_function(fun = dnorm,
                args = list(mean = 4, sd = 0.4),
                aes(color = 'Daytime Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Hatchery Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique HNC Fish'),
                geom = 'blank') +
  scale_color_brewer(palette = "Spectral") +
  # scale_color_grey(start = 0.9,
  #                  end = 0.3) +
  geom_vline(xintercept = 4,
             linetype = 2) +
  coord_cartesian(xlim = my_xlim,
                  ylim = my_ylim) +
  theme_pubr(legend = "bottom") +
  guides(color = guide_legend(nrow = 3,
                              title.position = 'top')) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank()) +
  labs(x = 'Escapement',
       color = 'Fish Type',
       title = '   Daytime Fish')

@@

examp_p = ggarrange(p1, p2a, p3a, p4,
                    labels = 'AUTO',
                    ncol = 1,
                    nrow = 4,
                    common.legend = T,
                    legend = 'bottom')

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
- AckermanLibrary.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

Ryan and Kevin decided to not go down this path. One quick comparison is total escapement vs. window counts; which shows up in one of our tables, and we discuss it briefly in the results. The other is figure 4, showing nighttime passage and re-ascension rates, and how they don't match. -->

@@

```{r stadem-examp-fig, fig.width=3.5, fig.asp = 2, fig.cap = "Schematic of how the STADEM model works. (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the re-ascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportions by origin (D), to estimate unique fish by origin."}

@@

# choose appropriate color scale for color vs. black & white
p1 = ggplot(data.frame(x = c(-5, 8)),
            aes(x)) +
  stat_function(fun = dnorm,
                args = list(mean = 4, sd = 0.4),
                aes(color = 'Daytime Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Hatchery Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique HNC Fish'),
                geom = 'blank') +
  scale_color_brewer(palette = "Spectral") +
  # scale_color_grey(start = 0.9,
  #                  end = 0.3) +
  geom_vline(xintercept = 4,
             linetype = 2) +
  theme_pubr(legend = "bottom") +
  guides(color = guide_legend(nrow = 3,
                              title.position = 'top')) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank()) +
  labs(x = 'Escapement',
       color = 'Fish Type',
       title = '   Daytime Fish') +
  coord_cartesian(xlim = my_xlim,
                  ylim = my_ylim)
# coord_fixed(ratio = 4,
  #             xlim = my_xlim,
  #             ylim = my_ylim)

@@

examp_p = ggarrange(p1, p2a, p3a, p4,
                    labels = 'AUTO',
                    ncol = 1,
                    nrow = 4,
                    align = 'v',
                    common.legend = T,
                    legend = 'bottom')
